Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
last batch of typos

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq. Their development involves
around 10,000 lines of Agda code.
\citet{martin-lof-a-la-coq} transports the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system. Their development is larger and has around
30,000 lines of Coq code.
% \scw{How big are these proofs?}

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. This metatheory is an
extensional type theory with features such as dependent functions,
inductive types, partial types, and a full universe hierarchy. They construct a PER
model in Coq to show the logical consistency of their language. Their
development has been further extended with features such as
intersection types, union types, and quotient types. The extensive
coverage of features result in a Coq development with around
330,000 lines of code.
\citet{nbeincoq}
mechanize the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe. Their development has
around 20,000 lines of Coq code.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further show it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion. Their encoding of a countable
universe hierarchy relies on impredicativity, similar to our
development.
% \scw{How did the Nuprl
% proof include a full hierarchy? Did they use impredicativity?}
% \yl{commented out because I think this paragraph is too subjective}
% However, instead of explaining the inductive
% logical relation as a convoluted workaround to the lack of
% induction-recursion in Coq, we give a self-contained explanation of
% our logical relation.

@@

\citet{barras2010sets, Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous efforts, which primarily focus on predicative
type theory and more direct reducibility models,
\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
$CC^\omega$, a system that incorporates a predicative universe on top
of the impredicative sort in Calculus of Constructions. We choose to
focus on a syntactic term model so we do not have to take the extra step
of mechanizing mathematical objects such as sets and domains.

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} proves the correctness of a type checker
for the Polymorphic, Cumulative Calculus of Inductive Constructions (PCUIC),
Coq's core calculus, assuming the strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type soundness of the type
system. Compared to properties such as consistency, normalization, and
parametricity, the most difficult part for syntactic type soundness is the
consistency of definitional (not propositional) equality. In their setting,
the result follows from the confluence of parallel reduction.

@@

The short consistency proof we present achieves the goal of
demonstrating the technique of proof by logical relation for dependently typed
languages. However, what remains unanswered is what makes our development
significantly shorter compared to others. Are we proving simpler results, for
smaller languages, are we making more use of automation, or is our proof
technique genuinely more efficient?
% focus on the following question: why is our proof,
% even with its extension to the existence $\beta\eta$-normal form, so much
% shorter than the other mechanized results from \citet{decagda,
% nbeincoq, martin-lof-a-la-coq}?

@@

First, the metatheoretic properties that we prove are indeed
simpler. Compared to Core Nuprl~\citep{anand2014towards}, our system
lacks extensionality, which would require a relational model to
justify consistency. Compared to the systems from
\citet{decagda,nbeincoq, martin-lof-a-la-coq}, the conversion rule for
\lang{} is untyped and therefore we do not
need a Kripke-style relational model to prove $\Pi$-injectivity
among other properties. Furthermore, we prove the existence of normal
forms, which induces a simple \emph{normalize-and-compare}
procedure for type
conversion~\citet{pierce2004advanced}. \citet{nbeincoq, decagda}, on
the other hand, need
to show how their algorithmic conversion procedure is sound and complete
with respect to their respective declarative equational theory.
\scw{I'm getting confused by this paragraph. Does this reorganization sense:
  Our language is simpler than Nuprl, because it doesn't have extensional
  equality. It is simpler than Agda, because it doesn't have type-directed
  equality. Both of these cases require the definition of a binary logical
  relation, that defines a notion of semantic equality between terms. This
  relation justifies the injectivity of $\Pi$ types and justify the validity
  of $\eta$-conversion among other properties.}  \scw{Furthermore our proof is
  also simpler because we don't need prove the correctness of the NBE
  algorithm, which is used to show the decidability of Agda's type-directed
  equivalence. Therefore, we don't need to define this algorithm and show that
  it is sound and complete with respect to the type-directed
  equality. Instead, to show the decidability of our untyped equivalence, we
  need only show that terms have $\beta\eta$ normal forms. }
\yl{Makes sense. Though Abel's work doesn't use nbe but a recursive
  binary algorithm. Rewrote the paragraph above and commented out the original}

Second, the definition of our logical relation does
contribute to a more concise proof.
In \rref{I-Red, I-Bool}, we choose parallel reduction, a full
reduction relation, to close over our semantic interpretation of types
and terms. Parallel reduction is non-deterministic, but it satisfies
useful structural properties such as congruence
(Lemma~\ref{lemma:parcong}) and the diamond property
(Lemma~\ref{lemma:pardiamond}). We pay the price of using a
non-deterministic reduction relation when we want to prove that our
logical relation is a partial function; because of \rref{I-Red}, we
can have $[[A => B0]]$ and $[[A => B1]]$, where $[[B0]]$ and $[[B1]]$
each have their separate interpretations that we have to prove to be
equal. Fortunately, this complexity is reconciled by the
diamond property, which is easy to derive syntactically.
\citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. The use of a deterministic reduction relation
makes the functionality of a logical relation trivial to prove, but
the downside is that weak head reduction (denoted by $\leadsto^{wh}$)
or leftmost-outermost reduction (denoted by $\leadsto^{lo}$) does
not satisfy even the weaker substitution property
(Lemma~\ref{lemma:parsubst}).
With this alternative formulation, we would have to prove how the
deterministic reduction relates to a full non-deterministic reduction
relation to prove the fundamental theorem. This would amount
to proving the factorization theorem of the deterministic reduction
relation with respect to full non-deterministic reduction. Instead,
using parallel reduction in \rref{I-Red} allows us to delay the
factorization property until we need to justify the use leftmost-outermost
reduction as a deterministic reduction strategy for normalization.

@@

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. \scw{Need to define Kripke-style. Also the other
proofs need Kripke style because they are defining typed relations, not untyped
relations. } \yl{I wonder if
we can just assume some more technical knowledge from the readers in
this section.}
However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available because it
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.\scw{confluence depends on Pi injectivity? I thought
it was only needed for subject reduction}\yl{it's
transitive. Confluence depends on subject reduction, which in turn
depends on pi injectivity. Maybe it's worth spelling out the details}
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover confluence property independently
from the logical relation. \citet{siles2012pure} generalize the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Type
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, thus allowing subject reduction to be proven
independently from confluence. By adopting these techniques that allow
us to derive confluence early even for systems with type-directed
reduction, we believe our proof technique can significantly shorten
the existing logical relation proofs for systems with typed
judgmental. We leave that as part of our future work.
% Therefore, we believe that even in a
% system where type-directed reduction is required (e.g. a system with
% the unit $\eta$-law) in the logical relation, the proof can still be carried
% out in a structure similar to the one we have presented.
% \scw{Not sure that I understand this paragraph}
% \yl{Reworded slightly to emphasize it's future work that we haven't
%   done and it is speculative}

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we show a short and mechanized proof of consistency
through proof by logical relation for a fully dependently typed
language with a full universe hierarchy, an intensional identity type,
and large elimination. We show the extensibility of our approach by
proving the existence of $\beta\eta$-normal forms with only small
and mechanical changes to our proof development. Our Coq mechanization
leverages existing Coq libraries for reasoning about metatheory and
general-purpose automation, allowing us to significantly reduce the
verbosity typically associated with mechanized proofs and recover a
declarative proof style that is usually only available in pen and
paper. Our development shows that proof by logical relation for
dependent types should be more accessible since it does not always
require months of effort to implement. We hope our proof can inspire
researchers to more actively mechanize results such as consistency,
normalization for dependent type theory.

>>>>

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and a typed judgmental equality that includes the function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq. Their development involves
around 10,000 lines of Agda code.
\citet{martin-lof-a-la-coq} transports the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system. Their development is larger and has around
30,000 lines of Coq code.
% \scw{How big are these proofs?}

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. This metatheory is an
extensional type theory with features such as dependent functions,
inductive types, partial types, and a full universe hierarchy. They construct a PER
model in Coq to show the logical consistency of their language. Their
development has been further extended with features such as
intersection types, union types, and quotient types. The extensive
coverage of features result in a Coq development with around
330,000 lines of code.
\citet{nbeincoq}
mechanize the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe. Their development has
around 20,000 lines of Coq code.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further show it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion. Their encoding of a countable
universe hierarchy relies on impredicativity, similar to our
development.

@@

\citet{barras2010sets, Wang2013SemanticsOI} assign
set-theoretic semantics to dependent type theory in Coq. Unlike the
previous efforts, which primarily focus on predicative
type theory and more direct reducibility models,
\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
$CC^\omega$, a system that incorporates a predicative universe on top
of the impredicative sort in the Calculus of Constructions. We choose to
focus on a syntactic term model so we do not have to take the extra step
of mechanizing mathematical objects such as sets and domains.

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
for the Predicative, Cumulative Calculus of Inductive Constructions (PCUIC),
Coq's core calculus, assuming the strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type soundness of the type
system. Because System D includes nontermination, they proved
the consistency of definitional equality from the confluence of 
parallel reduction.

@@

The short consistency proof we present achieves the goal of
demonstrating the technique of proof by logical relation for dependently typed
languages. However, what remains unanswered is what makes our development
significantly shorter compared to others. Are we proving simpler results for
smaller languages, are we making more use of automation, or is our proof
technique genuinely more efficient?
% focus on the following question: why is our proof,
% even with its extension to the existence $\beta\eta$-normal form, so much
% shorter than the other mechanized results from \citet{decagda,
% nbeincoq, martin-lof-a-la-coq}?

@@

First, the metatheoretic properties that we prove are indeed
simpler. Compared to Core Nuprl~\citep{anand2014towards}, our system
lacks extensionality, which would require a relational model to
justify consistency. Compared to the systems from
\citet{decagda, nbeincoq, martin-lof-a-la-coq}, the conversion rule for
\lang{} is untyped and therefore we do not
need a Kripke-style relational model to prove $\Pi$-injectivity
among other properties. Furthermore, we prove the existence of normal
forms, which induces a simple \emph{normalize-and-compare}
procedure for type
conversion~\cite{pierce2004advanced}. \citet{nbeincoq, decagda}, on
the other hand, need
to show how their algorithmic conversion procedure is sound and complete
with respect to their respective declarative equational theory.
\scw{I'm getting confused by this paragraph. Does this reorganization sense:
  Our language is simpler than Nuprl, because it doesn't have extensional
  equality. It is simpler than Agda, because it doesn't have type-directed
  equality. Both of these cases require the definition of a binary logical
  relation, that defines a notion of semantic equality between terms. This
  relation justifies the injectivity of $\Pi$ types and justify the validity
  of $\eta$-conversion among other properties.}  \scw{Furthermore our proof is
  also simpler because we don't need prove the correctness of the NBE
  algorithm, which is used to show the decidability of Agda's type-directed
  equivalence. Therefore, we don't need to define this algorithm and show that
  it is sound and complete with respect to the type-directed
  equality. Instead, to show the decidability of our untyped equivalence, we
  need only show that terms have $\beta\eta$ normal forms. }
\yl{Makes sense. Though Abel's work doesn't use nbe but a recursive
  binary algorithm. Rewrote the paragraph above and commented out the original}

Second, the definition of our logical relation does
contribute to a more concise proof.
In \rref{I-Red, I-Bool}, we choose parallel reduction, a full
reduction relation, to close over our semantic interpretation of types
and terms. Parallel reduction is non-deterministic, but it satisfies
useful structural properties such as congruence
(Lemma~\ref{lemma:parcong}) and the diamond property
(Lemma~\ref{lemma:pardiamond}). We pay the price of using a
non-deterministic reduction relation when we want to prove that our
logical relation is a partial function; because of \rref{I-Red}, we
can have $[[A => B0]]$ and $[[A => B1]]$, where $[[B0]]$ and $[[B1]]$
each have their separate interpretations that we have to prove to be
equal. Fortunately, this complexity is reconciled by the
diamond property, which is easy to derive syntactically.

In contrast, \citet{decagda} and \citet{nbeincoq} employ a deterministic weak
head reduction relation. The use of a deterministic reduction relation makes
the functionality of a logical relation trivial to prove, but the downside is
that weak head reduction (denoted by $\leadsto^{wh}$) or leftmost-outermost
reduction (denoted by $\leadsto^{lo}$) does not satisfy even the weaker
substitution property (Lemma~\ref{lemma:parsubst}).  With this alternative
formulation, we would have to prove how the deterministic reduction relates to
a full non-deterministic reduction relation to prove the fundamental
theorem. This would amount to proving the factorization theorem of the
deterministic reduction relation with respect to full non-deterministic
reduction. Instead, using parallel reduction in \rref{I-Red} allows us to
delay the factorization property until we need to justify the use
leftmost-outermost reduction as a deterministic reduction strategy for
normalization.

@@

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. \scw{Need to define Kripke-style. Also the other
proofs need Kripke style because they are defining typed relations, not untyped
relations. } \yl{I wonder if
we can just assume some more technical knowledge from the readers in
this section.}
However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available because it
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.\scw{confluence depends on Pi injectivity? I thought
it was only needed for subject reduction}\yl{it's
transitive. Confluence depends on subject reduction, which in turn
depends on pi injectivity. Maybe it's worth spelling out the details}
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover the confluence property independently
from the logical relation. \citet{siles2012pure} generalize the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Type
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, thus allowing subject reduction to be proven
independently from confluence. By adopting these techniques that allow
us to derive confluence early even for systems with type-directed
reduction, we believe our proof technique can significantly shorten
the existing logical relation proofs for systems with typed
judgmental. We leave that as part of our future work.
% Therefore, we believe that even in a
% system where type-directed reduction is required (e.g. a system with
% the unit $\eta$-law) in the logical relation, the proof can still be carried
% out in a structure similar to the one we have presented.
% \scw{Not sure that I understand this paragraph}
% \yl{Reworded slightly to emphasize it's future work that we haven't
%   done and it is speculative}

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof of consistency through
proof by logical relations for a dependently typed language with a full
universe hierarchy, large eliminations, an intensional identity type, and
dependent eliminators. We show the extensibility of our approach by proving
the existence of $\beta\eta$-normal forms with only small and mechanical
changes to our proof development. Our Coq mechanization leverages existing Coq
libraries for reasoning about metatheory and for general purpose automation,
allowing us to significantly reduce the verbosity typically associated with
mechanized proofs. The result is a declarative proof style that is usually
only available on pen and paper. Our development shows that proofs by logical
relations for dependent types are accessible and do not
require months of effort to implement. We hope our proof can inspire
researchers to more frequently mechanize results, such as consistency and
normalization, for their dependent type theories.
